---
title: "Medical Statistician Presentation"
subtitle: "NTNU Interview Case" 
author: "Artem Shiryaev"
date: "2023-11-28"
output: beamer_presentation
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Research Problem

Vi skal genomföre en klinisk studie på nydiagnostiserte pasienter med myelomatose hvor vi ska sammenlign standardbehandling med en ny kombinasjon av experimentella medikamenter.


Fra tidigare studier kan forvente:

- Standard: 50$\%$ minimal residual disease (MRD)

Forskargruppe tror att vi kan oppnå en klinisk signifikant bedring om:

- Experimentell: 70$\%$ minimal residual disease (MRD)

Med andre ord, jo mindre kreft, desto bedre.

## Statistical Analys

Head of the Research team thinks: Data costs money,

Hvor mange patienter trenger studien vår, med $80\%$ power og signifikansnivå på 5 %?

## Resultat beroende data (paired/dependent data)
Beste fall: total 98 patienter, 49 vardera gruppe\
Verste fall: 102\


| Test          | Same  | Assumptions |
|:-------------:|:-----:|:----------:|
| Two Sample T  | 49 + 49  | Approx. Normal, iid |
|  Wilcoxon signed-rank Test |   51 + 51 | Non-parametric, iid, effect size $\approx$ 0.41|


## Resultat oberoende data (independent data)
Beste fall: total 182 patienter, 91 vardera gruppe\
Verste fall: 231\


| Test          | Same | Different | Assumptions |
|:-------------:|:-----:|:----------:|:------------:|
| Two Sample Z  |91 + 91|    -    | Normality, iid|
| Two Sample T  |   93 + 93 |  180 + 63 | Approx. Normal, iid |
| Mann-Whitney U  |   99 + 99 |  -    | Non-parametric, iid, effect size $\approx$ 0.41|
|Chi$^2$ ($df=90$)| 231 | -  | Z scores for df |





```{r, echo=TRUE, include=FALSE}
library(pwr)
library(pwrss)
library(ggplot2)
library(scales)


Prop_DrugA <- 0.5 
Prop_DrugB <- 0.7

Alpha <- 0.05 # Significance Level
Beta <- 0.8   # Power Level

# Difference between Two Proportions (z Test)
z_prop <- pwrss.z.2props(p1 = Prop_DrugA,
                         p2 = Prop_DrugB,
                         alternative = "not equal",
                         power = Beta,
                         alpha = Alpha
  
)


# Effect Size using Binomial Transform
h_prop <- 2*asin(sqrt(Prop_DrugA)) - 2*asin(sqrt(Prop_DrugB))

# Power calculation for two proportions (same sample sizes)
two_test <-pwr.2p.test(h=round(h_prop, digits = 2), 
                       sig.level = Alpha,
                       power = Beta,
                       alternative = "two.sided")


# Power Calculation for two proportions (different sizes)
diff_size <- pwr.2p2n.test(h=round(h_prop, digits = 2),
                           n1 = 180,
                           power =  Beta,
                           sig.level = Alpha,
                           alternative="two.sided")

chi2_tst <- pwr.chisq.test(w = (-1)*round(h_prop, digits = 2),
                           N = NULL,
                           df = (91-1)*(2-1),#(r-1)(c-1),
                           sig.level = 0.05,
                           power = 0.8
  
)



# Power calculations for t-tests of means (one sample, two samples and paired samples)
t_test_2means <- pwr.t.test(d = round(h_prop, digits = 2),
                          sig.level = Alpha,
                          power = Beta,
                          type = "two.sample") 



# Power calculations for two samples (different sizes) t-tests of means
t.test_diff_means <- pwr.t2n.test(n1 = 180,
                                  d = round(h_prop, digits = 2),
                                  sig.level = Alpha,
                                  power = Beta,
                                  alternative = "two.sided")


# Difference between Two Groups (Non-parametric Tests for Independent and Paired Samples)
np_prop <- pwrss.np.2groups(mu1 = round(h_prop, digits = 2),
                            mu2 = 0,
                            alternative = "not equal",
                            power = Beta,
                            alpha = Alpha,
                            distribution = "normal"
                            )

# Paired samples

paired_t <- pwrss.t.2means(mu1 = round(h_prop, digits = 2),
                           alternative = "not equal",
                           paired = TRUE,
                           paired.r = 0.5,
                           alpha = Alpha,
                           power = Beta
  
)


paired_np_2groups <- pwrss.np.2groups(mu1 = round(h_prop, digits = 2),
                            mu2 = 0,
                            paired = TRUE,
                            paired.r = 0.5,
                            alternative = "not equal",
                            power = Beta,
                            alpha = Alpha,
                            distribution = "normal"
                            )

```

## Graphical Results Z Test
```{r, echo = F}
# Power calculation for two proportions (same sample sizes)
plot(z_prop)
#plot.power.htest(z_prop, main = "Two proportions Z test (same sample sizes)")
```

<!-- ## Graphical Results Student t-test -->
<!-- ```{r, echo = F} -->
<!-- # Power calculations for t-tests of means (one sample, two samples and paired samples) -->
<!-- plot.power.htest(t_test_2means, main = "Two means (same sample sizes)") -->
<!-- ``` -->



<!-- ## Graphical Results Student t-test -->
<!-- ```{r, echo = F} -->
<!-- # Power calculations for two samples (different sizes) t-tests of means -->
<!-- plot.power.htest(t.test_diff_means , main = "Two means (different sample sizes)") -->
<!-- ``` -->



## Graphical Results Student t-test
```{r, echo = F}
# Power calculation for two proportions (same sample sizes)
plot.power.htest(two_test, main = "Two proportions (same sample sizes)")
```

## Graphical Results Student t-test
```{r, echo = F}
# Power Calculation for two proportions (different sizes)
plot.power.htest(diff_size, main = "Two proportions (different sample sizes)")
```



## Graphical Results Chi$^2$
```{r, echo = F}
# Power Calculation for two proportions (different sizes)
plot.power.htest(chi2_tst, main = "Chi^2 Power Calculation")
```


## Code

[Github link](https://github.com/ArtemShiryaev/NTNU_Statistics/blob/master/NTNU_Script.R)




## Analytical Calculation: Z-tests

\[H_0 : p_1 = p_2\]
\[H_1 : p_1 \neq p_2 \]

To determine the sample size needed for a study comparing two proportions use  two-sample proportion test. The formula is:
\[Z = \frac{(p_1 - p_2 )}{\sqrt{\frac{p_1 (1-p_1 )}{n_1 } + \frac{p_2 ( 1-p_2 )}{n_2 }}}\]
Simplified to, given \( n_1 = n_2 = n\):
\[ n = \frac{{(p_1(1 - p_1) + p_2(1 - p_2))(Z_\alpha + Z_\beta)^2}}{{(p_1 - p_2)^2}} \]

## Analytical Calculation: Z-tests

\( p_1 \) is the reduction in MRD for Drug A (50$\%$, or 0.5).\

\(  p_2 \) is the targeted reduction in MRD for Drug B (70$\%$, or 0.7).\

\(  Z_\alpha \) is the Z-score for a $5\%$ significance level (approximately 1.96).\

\(  Z_\beta \) is the Z-score for an 80$\%$ power (approximately 0.84). 



## Analytical Calculation: Z-tests

Substitute these values into the formula to calculate \( n \):

\[ n = \frac{{(0.5 \times 0.5 + 0.7 \times 0.3)(1.96 + 0.84)^2}}{{(0.5 - 0.7)^2}} \]

Now, calculate the result. 

\[ n = \frac{{(0.25 + 0.21)(2.8)^2}}{{0.04}} \]

\[ n = \frac{{0.46 \times 7.84}}{{0.04}} \]

\[ n = \frac{{3.6064}}{{0.04}} \]

\[ n = 90.16 \]
\[ n \approx 91\]

## Analytical Calculation: Student t-test


Start with the formula for the t-statistic in the two-sample proportion t-test:

\[ t = \frac{(\hat{p}_1 - \hat{p}_2 )}{\sqrt{\frac{\hat{p}_1 (1-\hat{p}_1 )}{n_1 } + \frac{\hat{p}_2 ( 1-\hat{p}_2 )}{n_2 }}} \]

For the t-test, the formula is:

\[ n = \frac{{(t_{\alpha/2, df} + t_\beta)^2 \times (p_1(1 - p_1) + p_2(1 - p_2))}}{{(p_1 - p_2)^2}} \]

\(\hat{p}_1\) and \(\hat{p}_2\) are the sample proportions for Group 1 and Group 2, 
 \( \hat{p}_1 = 0.5 \) (for Drug A),\
 \( \hat{p}_2 = 0.7 \) (for Drug B),\
 Degrees of freedom (\( df \)) for the t-distribution: \( df = 2n - 2 \),\
\( t_{\alpha/2, df} \) and \( t_\beta \) from the t-distribution tables.


## Analytical Calculation: Student t-test

\[ n = \frac{{(1.97 + 0.85)^2 \times (0.5(1 - 0.5) + 0.7(1 - 0.7))}}{{(0.5 - 0.7)^2}} \]

\[ n \approx \frac{{(2.82)^2 \times (0.25 + 0.21)}}{{0.04}} \]

\[ n \approx \frac{{(7.9524) \times (0.46)}}{{0.04}} \]

\[ n \approx \frac{{3.656704}}{{0.04}} \]

\[ n \approx 91.42 \]
\[ n \approx 92 \]






